network_plot

### 2024-02-08 R/MASCARA_Tutorial.Rmd
24_02_08

User must supply a dataframe of filtered + transformed data plus associated metadata.

@@

meta <- cbind.data.frame(rep(c(1,-1), each = 12), rep(c(1:4), each = 3), rep(c(-1:-4,1:4), each = 3))
colnames(meta) <- c("growth_condition","time","interaction")
meta <- as.data.frame(apply(meta,2,factor))  #9

Experiment_responders <- 30

set.seed(3)
# rnd_seed <- abs(rnorm(3, mean = 0.7, sd = 0.5))

baits <- paste0("X_",c(1985:1988))   #501:510
spikes <- paste0("X_",c(1989:2000))

sim_data <- Create_Core_DEV_2(nreps = 3, meta = meta, plot = TRUE, 
                        EffectSize = c(X_a_ab = 1.8, time = 1, E = 10, mu = 1, Struc_E = 1), 
                        struc_resid = TRUE, e_sigma = c(1.5,0.8,0.0), a_sigma = c(1,0.7,0.6,0.1),
                        b_sigma = c(1,0.9), irr_spikes = TRUE, SCORE_SEED = 12, 
                        Experiment_responders = Experiment_responders, struc_seed = 127)

@@

```{r}
ar <- ASCA(sim_data[[1]], m = meta, Baits = baits, spikes = spikes, ncands = ncands, distance_calc = FALSE, Return_Model = FALSE)
```

```{r}
#classic
MASCARA_2 <- MASCARA4_test_2(ar[[3]], baits, spikes = spikes)
which(rownames(MASCARA_2[[2]]) %in% spikes)
```

>>>>

User must supply a dataframe of filtered + transformed data plus associated metadata, with samples in the rows and genes in the columns.

```{r}
meta <- cbind.data.frame(rep(c(1,-1), each = 12), rep(c(1:4), each = 3), rep(c(-1:-4,1:4), each = 3))
colnames(meta) <- c("growth_condition","time","interaction")
meta <- as.data.frame(apply(meta,2,factor))  
knitr::kable(head(meta))
```

Data is assumed to be approximately normally distributed:

```{r setting simulation parameters}
Experiment_responders <- 30

set.seed(3)

baits <- paste0("X_",c(1985:1988))  
spikes <- paste0("X_",c(1989:2000))

sim_data <- Create_Core_DEV_2(nreps = 3, meta = meta, plot = TRUE, 
                        EffectSize = c(X_a_ab = 1.8, time = 1, E = 10, mu = 1, Struc_E = 1), 
                        struc_resid = TRUE, e_sigma = c(1.5,0.8,0.0), a_sigma = c(1,0.7,0.6,0.1),
                        b_sigma = c(1,0.9), irr_spikes = TRUE, SCORE_SEED = 12, 
                        Experiment_responders = Experiment_responders, struc_seed = 127)[[1]] 
hist(sim_data)
```

We calculate an ASCA model on the data to partition the experimental and residual variance:

```{r}
ar <- ASCA(sim_data, m = meta, Baits = baits, spikes = spikes, ncands = ncands, distance_calc = FALSE, Return_Model = FALSE)
resids <- ar[[3]]
```

Before continuing it is a good idea to check that the set of baits are indeed correlated in the residual variance:

```{r}
cor(resids[,which(colnames(resids) %in% baits)])
```
In this case they are, in the case of real data we expect these values to be somewhat lower (values ~ 0.6 are reason enough to continue).

Another sanity check at this stage is to look at the baits residual variance against eachother:

@@

plot(resids[,baits[1]], resids[,baits[2]])

@@

We can see above that there is a relationship between the examples.

From here we take the residuals and calculate a PLS2 model with a predefined set of baits using the SIMPLS algorithm:

spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
                          resids[,which(colnames(resids) %in% baits)], ncomp = 2)

Selecting features with $\mathbf{\bar{q}}$ target projection:

```{r}
q_bar <- colMeans(spls_res$Yloadings)
R <- spls_res$projection
R_TP <- (dot(t(R),q_bar)/dot(q_bar,q_bar))

Candidates <- as.data.frame(R_TP[order(R_TP, decreasing = TRUE)])
```

Now we have our ordered candidate list, top of the list represents features that have the strongest positive association with the baits.

### 2024-02-08 R/Paper_markdown.Rmd
24_02_08

# - name: Fred T.G. White
#     email: f.t.g.white@uva.nl
#     affiliation: 
#       - University of Amsterdam
#       - Another example
#     corresponding: alice@example.com
#   - name: Bob Security
#     email: bob@example.com
#     affiliation: 
#       - Another University
#       - Some Institute of Technology
# address:
#   - code: Univeristy of Amsterdam
#     address: Department 1, Street, City, State, Zip
#   - code: Another example
#     address: Department 2, Street, City, State, Zip
```

@@

```{r nettype, fig.cap= c("Pathway types exemplified, POI; pathway of interest, RP; related pathway, URP; unrelated pathway. Panel A represents hypothetical biological pathways  that are all affected by the experimental design and thus all contain DE genes, here the edges represent real connections between the nodes/genes. Panel B highlights the problem with detecting these pathways when dealing with data from designed experiments, edges represent putative connections defined by coexpression analysis; black is desired (within network detection), grey non-desired.")}

@@

1. Initialise \( \mathbf{\tilde{Y}}_0 \) to \( \mathbf{\tilde{Y}} \) and \( \mathbf{B}_0 \) to \( \mathbf{B} \).
2. Iteratively perform the following steps for \( k = 0, 1, 2, \ldots \):
   a. Compute the singular value decomposition (SVD) of \( \mathbf{\tilde{Y}}_{k}^{T} \mathbf{B}_{k} \) to obtain the left and right singular vectors; weights \( \mathbf{u}_k \) and \( \mathbf{v}_k \) respectively.
   b. Project \( \mathbf{\tilde{Y}}_{k} \) and \( \mathbf{B}_{k} \) onto the first left and right singular vectors (\( \mathbf{u}_k \), \( \mathbf{v}_k \)) to obtain the scores \( \bm{t}_k = \mathbf{\tilde{Y}}_{k} \mathbf{u}_k \) and \( \bm{w}_k = \mathbf{B}_{k} \mathbf{v}_k \).

Under the conditions of orthogonal scores; $\mathbf{t}_{k}^{T}\mathbf{t}_{k} = 0$ and normalised weights; $\mathbf{u}_{k}^T\mathbf{u}_{k{}} = 1$ and $\mathbf{v}_{k}^T\mathbf{v}_{k{}} = 1$.

PLS2 aims to maximize the covariance between the scores \( \bm{t}_k \) and \( \bm{w}_k \) at each iteration. In the following simulations and application the number of components calculated by PLS2 is set to 2 to create an interpretable model.

@@

Where k=1...K represents the number of components in the respective models, j indexes a gene,  $H$ is substituted for the loadings $\mathbf{P}_{\alpha + \alpha\beta}^T$ or weights $\mathbf{U}$ (for ASCA+ and PLS respectively) and VE is the variance explained, where $VE_{k}$ is the amount of variance explained by component k and $VE_{total}$ is the variance explained by the whole model. In ASCA+ we calculate the VIP scores in the latent space of \(\bm{Y}_{\alpha + \alpha\beta}\) as described in @thielASCAAPCAExtensions2017. In PLS we use the VE from the latent space of \(\bm{B}\) using the weights ($\mathbf{U}$) from the latent space of \(\bm{Y}\).

>>>>

# - name: Fred T.G. White
#     email: f.t.g.white@uva.nl
#     affiliation: 
#       - University of Amsterdam
#       - Another example
#     corresponding: alice@example.com
#   - name: Bob Security
#     email: bob@example.com
#     affiliation: 
#       - Another University
#       - Some Institute of Technology
# address:
#   - code: Univeristy of Amsterdam
#     address: Department 1, Street, City, State, Zip
#   - code: Another example
#     address: Department 2, Street, City, State, Zip

## CHANGE THIS   
#"one comment already, since I am looking at it right now. In the readme “~/MASCARA/R” , the repo (clone) most likely doesn’t live there. Maybe start from ‘clone this repo’ and then let people set the wd to where they put the repo. Or put the snippet from the readme into an r script that you let them open in rstudio to make sure they’re in the right place" AHB
```

@@

```{r nettype, fig.cap= c("Pathway types exemplified, POI; pathway of interest, RP; related pathway, URP; unrelated pathway. Panel A represents hypothetical biological pathways  that are all affected by the experimental design and thus all contain DE genes, here the edges represent real connections between the nodes/genes. Panel B highlights the problem with detecting these pathways when dealing with data from designed experiments, here edges represent putative connections defined by coexpression analysis; black is desired (within network detection), grey non-desired.")}

@@

<!-- 1. Initialise \( \mathbf{\tilde{Y}}_0 \) to \( \mathbf{\tilde{Y}} \) and \( \mathbf{B}_0 \) to \( \mathbf{B} \). -->
<!-- 2. Iteratively perform the following steps for \( k = 0, 1, 2, \ldots \): -->
<!--    a. Compute the singular value decomposition (SVD) of \( \mathbf{\tilde{Y}}_{k}^{T} \mathbf{B}_{k} \) to obtain the left and right singular vectors; weights \( \mathbf{u}_k \) and \( \mathbf{v}_k \) respectively. -->
<!--    b. Project \( \mathbf{\tilde{Y}}_{k} \) and \( \mathbf{B}_{k} \) onto the first left and right singular vectors (\( \mathbf{u}_k \), \( \mathbf{v}_k \)) to obtain the scores \( \bm{t}_k = \mathbf{\tilde{Y}}_{k} \mathbf{u}_k \) and \( \bm{w}_k = \mathbf{B}_{k} \mathbf{v}_k \). -->

1. For the $1^{st}$ component:
   a. Compute the left singular vector $\mathbf{u}_1$ of $\mathbf{\tilde{Y}}_{k}^{T} \mathbf{B}_{k}$
   b. $\mathbf{t}_1 = \mathbf{\tilde{Y}}\mathbf{u}; \mathbf{T} = \left [\mathbf{t}_1\right]$
   c. $\mathbf{p}_1 = \frac{\mathbf{\tilde{Y}}^T\mathbf{t}^t}{\mathbf{t}^t\mathbf{t}}$
2. For the following components:
   a. Compute the left singular vector $\mathbf{u}_k$ of $(\mathbf{I} - \mathbf{P}(\mathbf{P}^T\mathbf{P})^{-1}\mathbf{P})\mathbf{\tilde{Y}}_{k}^{T} \mathbf{B}_{k}$
   b. $\mathbf{t}_k = \mathbf{\tilde{Y}}\mathbf{u}_k; \mathbf{T} = \left [\mathbf{T}|\mathbf{t}_k\right]$
   c. $\mathbf{p}_k = \frac{\mathbf{\tilde{Y}}^T\mathbf{t}_{k}^t}{\mathbf{t}_{k}^t\mathbf{t}_{k}}; \mathbf{P} = \left [\mathbf{P}|\mathbf{p}_k\right]$

Under the conditions of orthogonal scores; $\mathbf{t}_{k}^{T}\mathbf{t}_{k} = 0$ and normalised weights; $\mathbf{u}_{k}^T\mathbf{u}_{k{}} = 1$ . Loadings for B can be obtained with a least squares step $\mathbf{Q} = (\mathbf{T}^T\mathbf{T})^{-1}\mathbf{T}^T\mathbf{B}$.

In the following simulations and application the number of components calculated by PLS2 is set to 2 to create an interpretable model.

<!-- and $\mathbf{v}_{k}^T\mathbf{v}_{k{}} = 1$ -->
<!-- PLS2 aims to maximize the covariance between the scores \( \bm{t}_k \) and \( \bm{w}_k \) at each iteration. -->

@@

Where k=1...K represents the number of components in the respective models, j indexes a gene,  $H$ is substituted for the loadings $\mathbf{P}_{\alpha + \alpha\beta}^T$ or weights $\mathbf{U}$ (for ASCA+ and PLS respectively) and VE is the variance explained, where $VE_{k}$ is the amount of variance explained by component k and $VE_{total}$ is the variance explained by the whole model. In ASCA+ we calculate the VIP scores in the latent space of \(\bm{Y}_{\alpha + \alpha\beta}\) as described in @thielASCAAPCAExtensions2017. In PLS we use the VE from the latent space of \(\bm{B}\) using the weights ($\mathbf{U}$) from the latent space of the predictor matrix \(\bm{Y}\).

### 2024-02-08 R/run_first/2.plot_pathway_net_example.Rmd
24_02_08

@@

png("../networktypes.png", height = 3600, width = 4800, res = 700)
realplot + badplot + plot_annotation("Pathway Types", tag_levels = 'A') + plot_layout(guides = "collect")
dev.off()

>>>>

```{r}
#sparsify -1 connections

# adj[which(adj == -1)] <- adj[which(adj == -1)] * sample(c(1,0,0,0))

# adj[which(adj == -1, arr.ind = T)] <- adj[which(adj == -1,arr.ind = T)] * sample(c(1,0,0,0))

# W <- matrix(sample(c(1,1,0,0,0),length(baits) * length(spikes), replace = T), nrow = length(spikes))
# 
# 
# adj[which(rownames(adj) %in% spikes),which(colnames(adj) %in% baits)] <- adj[which(rownames(adj) %in% spikes),which(colnames(adj) %in% baits)] * W

@@

# png("../networktypes.png", height = 3600, width = 4800, res = 700)
realplot + badplot + plot_annotation("Pathway Types", tag_levels = 'A') + plot_layout(guides = "collect")
# dev.off()
